Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-11-12 Dail_Mad_ext.tex
up to sec2.4

Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The exponential growth model is the simplest: 
$f(N_{i,t-1}) = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent, exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.
Because a single Poisson distribution controls the
distribution of $N_{i,t}$ in each of these models, the discrete
convolution used by \citet{dail_madsen:2011} to construct the likelihood is not required,
speeding up processing time.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0
because all contributions to population growth are assumed to be
local.  We therefore generalized these models to
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

\citet{dail_madsen:2011} suggested two distributions for modeling
initial abundance: Poisson and negative binomial. We have
extended their model to include another distribution, the zero-inflated
Poisson. This could be useful when, for example, one is modeling
the abundance of several species with the same set of point
count surveys for all species, but some survey sites are
outside the range of some of the species. The distribution
of initial abundances can be represented as:
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\Lambda &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
where $\psi$ represents the proportion of extra zeros.

These models allow three sources of zero counts by observers: a
species was at a site but not detected; the site was within the
species' range but no individuals occured at that site in that
year; and the site was outside the species' range. Furthermore,
detection, abundance, and zero-inflation can be modeled separately as
functions of different (or the same) covariates. For example, detection of a
species might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.  
%This approach combines elements of occupancy modeling
%\citep{mackenzie_etal:2006} and abundance modeling.
%, and could prove useful
%for distinguishing species' fundamental and realized niches
%(Hutchinson 1957).

@@

The zero-inflated Poisson distribution can be applied to recruitment and 
population growth terms as well as initial abundance. For example, 
the recruitment term of the constant-recruitment DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{i,t} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Pois}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as
zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case,
$\psi$ can also be
modeled as varying over space). In the latter case, it makes
sense
for the model to "remember" which sites are inside the range of
the
species between the initial abundance stage and each time step
of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
models
(without immigration) should have no effect when initial
abundance is
also zero-inflated unless $\psi$ is time-varying (because these
dynamic
models cannot recover from zero abundance).
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

>>>>

Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The exponential growth model is the simplest: 
$f(N_{i,t-1}) = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent, exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the Ricker-logistic population growth model
\citep{ricker:1954}. Another option is a modified Gompertz-logistic
density-dependent model \citep{hart_gotelli:2011}:
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

@@

One limitation of the geometric-recruitment, exponential,
Ricker-logistic, and Gompertz-logistic versions of the DM models all
%share a common feature (or bug): once the 
is that they include no mechanism for a local population to recover
after extinction. %at a site reaches 0, it must remain at 0
%because all contributions to population growth are assumed to be
%local.  
However, these models can be generalized 
%We therefore generalized these models 
to include both internal and external (immigration) contributions
to population growth. For example, an exponential plus immigration
model is:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

%\citet{dail_madsen:2011} suggested two distributions for modeling
%initial abundance: Poisson and negative binomial. We have
%extended their model to include another distribution, the zero-inflated
%Poisson. This could be useful when, for example, one is modeling
%the abundance of several species with the same set of point
%count surveys for all species, but some survey sites are
%outside the range of some of the species. 
The negative binomial distribution used in $N$-mixture models can have
a very long right tail, and its mean to variance ratio may not be
ideal in all cases. An alternative is to consider zero-inflated
distributions, such as the zero-inflated Poisson:
%The distribution
%of initial abundances can be represented as:
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
\mbox{Pois}(0) &\; \text{with probability} \; \psi \\
\mbox{Pois}(\Lambda) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
where $\psi$ represents the proportion of extra zeros.

%These models allow three sources of zero counts by observers: a
%species was at a site but not detected; the site was within the
%species' range but no individuals occured at that site in that
%year; and the site was outside the species' range. Furthermore,
Under this formulation, 
detection, abundance, and zero-inflation can be modeled separately as
functions of covariates. For example, detection of a
species might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.  
%This approach combines elements of occupancy modeling
%\citep{mackenzie_etal:2006} and abundance modeling.
%, and could prove useful
%for distinguishing species' fundamental and realized niches
%(Hutchinson 1957).

@@

The zero-inflated Poisson distribution can be applied to recruitment and 
population growth terms as well as initial abundance. For example, 
the recruitment term of the constant-recruitment DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{i,t} \sim \left\{
\begin{aligned}
\mathrm{Pois}(0) &\; \text{with probability} \; \psi \\
\mathrm{Pois}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as
zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case,
$\psi$ can also be
modeled as varying over space). In the latter case, it makes
sense
for the model to "remember" which sites are inside the range of
the
species between the initial abundance stage and each time step
of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
models
(without immigration) should have no effect when initial
abundance is
also zero-inflated unless $\psi$ is time-varying (because these
dynamic
models cannot recover from zero abundance).
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)
